The Symmetry of Partner Modelling

### 2016-02-23 mutual-modeling.tex
Fixed grammar errors pointed by editors

Collaborative learning has often been associated to the construction of  a
shared   understanding of the situation at hand. The psycholinguistics
mechanisms at work while establishing common grounds are the object of
scientific controversy. We postulate that collaborative tasks require some
level of mutual modelling, \ie that each partner needs some model of what
the other partners know/want/intend at a given time.  We use the term ``some
model'' to stress the fact that this model is not necessarily detailed or
complete, but that we acquire some representations of the persons we
interact with. The question we address is: Does the quality of the partner
model depend upon the modeler's ability to represent his or her partner?
Upon the modelee's ability to make his state clear to the modeler? Or rather upon
the quality of their interactions? We address this question by comparing
the respective accuracies of the models built by different team
members. We therefore report on 5 experiments on collaborative problem solving
or collaborative learning that vary in terms of tasks (how important it is
to build an accurate model) and settings (how difficult it is to build an
accurate model). In 4 studies, the accuracy of the model that \A built about
\B was correlated with the accuracy of the model that \B built about \A,  which
seems to imply that the quality of interactions matters more than individual
abilities when building mutual models.  However, these finding do not
rule out the fact that individual abilities also participate in the quality of
modelling process.

@@

From its inception, Computer Supported Collaborative Learning (CSCL)
research has been following the suggestion by \citet{roschelle1995construction}
that collaborative learning has something to do with the process of constructing
and maintaining a \emph{shared understanding} of the task at hand. Building a
shared/mutual understanding refers to the upper class of collaborative learning
situations, those in which students should build upon each other's understanding
to refine their own understanding: what is expected to produce learning is not
the mere fact that two students build the same understanding but the cognitive
effort they have to engage to build this shared
understanding~\citep{schwartz1995emergence}. This effort can be observed by the
frequency of rich interactions, \ie interactions whose occurrence has been
related to learning: (self-) explanations in cognitive
science~\citep{chi1989self, webb1991task}, conflict resolution in
socio-cognitive theories~\citep{doise1975social}, mutual
regulation~\citep{blaye1995collaborative} in a Vygostkian perspective, etc. The
construction of a shared understanding has been investigated for several decades
in psycholinguistics, under the  notion of
\emph{grounding}~\citep{clark1986referring}. However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned in the learning
sciences. Grounding mechanisms are appropriate to explain conversational
events, such as referential failures in short dialogue episodes, but they do hardly
predict deeper phenomena such as \emph{conceptual change} (\ie the acquisition,
acceptation and integration of a new belief into one's mental model) over longer
sessions~\citep{dillenbourg2006sharing}. The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:

@@

In summary, the controversy around the cognitive depth of shared understanding
pertains to psycholinguistic, which investigates natural conversations. The
situation is different in collaborative learning; tasks are actually designed
for requiring a deeper negotiation of meaning.  The expertise of CSCL is to design
collaborative situations that create interdependence, avoid group think, and
allows students to detect any ``illusion of shared
understanding''~\citet{cherubini2005grounding}. Our question is hence not
anymore ``do peers build a shared
understanding ?'', but rather ``whenever peers have to build a shared understanding,
how do they achieve it ?''.  This question addresses the mechanisms of grounding, in
which the basic sequence is: make a proposition, detect misunderstanding and, if
any, repair them.  This paper focuses on the middle part, the detection of
misunderstandings.  Detecting a misunderstanding means that the emitter of a
message identifies a mismatch between his communicative intention and the way
his message is understood by his or her partner. Detecting one peer's
misunderstanding is investigated hereafter under the general umbrella of partner
modelling.

@@

    \item $Model(A,B, ``urgent")$: what does \A know about \B's understanding of \A's
        last utterance: does ``urgent'' means now, soon or ``not too late''?

@@

This mutuality allows us to address a fundamental question: \emph{does the
quality of partner modelling depend upon the cognitive skills of each partner
(some people being better in perceiving other's states) or does it result from
the quality of interactions among them ?} Behind his question, the reader may
perceive the long lasting debate between tenets of, respectively, the individual
and social views of human cognition.  The simple hypothesis is that when
individual are randomly paired for an experiment, there is no reason for which
their individual cognitive skills would correlate. Therefore, if it occurs that
the quality of the $Model(A,B)$ is correlated with $Model(B,A)$, one may infers
that this quality depends upon what \A and \B have built together while
interacting.

To answer this question, we went back to five previous studies that addressed
various other research questions, but in which the correlation between the
quality of these two models could be computed.  The studies we report do hence
not constitute a clean sequence of experiment to investigate mutual modelling but
the ad-hoc revisiting of previous experiments to explore a question that we had
then neglected. Some of these studies are about collaborative learning while
others are only collaborative problem solving, but the latter rely on rich tasks
that are similar to those we use in CSCL.

@@

Natural language becomes cumbersome when describing things such as ``the model
that \A builds about the model that \B builds about A''. Therefore, to define
hypotheses and report on experiments on mutual modelling, we use  the notation
\model{A}{B}{X} to denote ``\A knows that \B knows X''. This  notation is not
proposed as a formal theory of mutual modelling but as useful simplification for
communicating about mutual modelling. This notation does not mean \A has an
explicit, monolithic representation of \B: it must be understood as an
abstraction referring to complex socio-cognitive processes.  As explained in the
previous section, \A can be fragmented, multi-dimensional, etc. This notation
is neither presented as a computational model of mutual modelling, nor as some
universal formalism; its usefulness is internal to this paper.  Besides, we
refer to the \emph{degree of accuracy} of the model as \Model{A}{B}{X}. We
discuss in the next section the methodological difficulty in measuring this
accuracy.

@@

    \item Before any specific grounding action, there is generally a non-null
        probability that $X$ is mutually understood by \A and \B (\eg $X$ is
        part of \A's and \B's cultures, it is manifest to co-present subjects
        or simply there is not much space for misunderstanding or disagreement
        about $X$). We simply could not collaborate without a certain level of
        initial grounds. We note the theoretical accuracy of \emph{initial
        grounds} $\mathcal{M}^{\circ}_{t_0}(A,B,X)$.
        \vspace{1em}

@@

We report on five studies conducted  by different researchers in different contexts
between 2000 and 2015. They do not form a consistent research strategy but
actually the fact that some trends emerge despite their diversity constitutes
indeed the richness of this line of work.

@@

How to interpret a correlation of 0.38 between \gModel{A}{B} and \gModel{B}{A}?
It is significant, which supports $\mathcal{H}_{2}$ and $\mathcal{H}_{3}$. It is
nonetheless far from 1 which implies we cannot discard the individual modelling
skills. We collect more evidence along the next studies.

@@

\paragraph{Rectangle question} We cannot address the relationship $\Delta_4$
between self and social accuracy here because we do not have an estimation of
self-accuracy: subjects describe their own emotions but we have no way to check if
they are correct. By measuring \gModel{A}{B} on  18 emotional labels,
we can however have a glimpse about $\Delta_5$: how \Model{A}{B}{X} varies
according to X.  Fig.~\ref{study3:deg_m_values} shows the range of modelling
errors: the difference between \gmodel{A}{B} and \gmodel{B}{B}, on a scale of 7,
is 0.3 in average for the emotion \emph{discasted}, and up to 1.9 for the
emotion \emph{calm}. This is probably specific to the variety of scales  ($SD=
0.5$ for \emph{discated} versus $SD=0.7$ for \emph{calm}), our point is not to
interpret this too far, but to show that there are large variations even within one
area  (perceiving emotions). These variations still question the existence of
a general aptitude to model others ( $\mathcal{H}_{1}$).

@@

\paragraph{Grounding criterion} A linear regression revealed a positive relation
between \gModel{A}{B} and the learning gain of the pair $\{A, B\}$ (calculated
as the average of the individual learning gains): $\beta= 0.401, p < 0.001, r2adj.
= 0.15$, large effect. This relationship was not significant in the previous
experiment which as been conducted with the same task (using the same task
should produce the same grounding criterion). This is probably due to the fact that
\gModel{A}{B} was influenced by the KAT.

@@

\paragraph{Symmetry question} We did not find an intra-pair correlation ($ICC =
0.05, ns$), which does not support $\mathcal{H}_{2}$ and $\mathcal{H}_{3}$. In
addition, the KAT supports the processing of modelling at the first level
($\mathcal{H}_{1}$) and not at the second level ($\mathcal{H}_{2}$). Hence the
fact that the KAT enhances  \gModel{A}{B}  brings additional supports to
$\mathcal{H}_{1}$. However, the analysis of verbal interactions find elements
that actually supports the three hypotheses.

@@

We found a symmetry of the mutual models on 4 studies out of 5. In study 4, this
only applies to the control group (having read the same text before), which
corresponds to the situation of the 3 first studies. Evidencing this symmetry
constitutes \textit{per se} an interesting result as we are not aware of earlier
studies who have established this relationship.
Still, the symmetry alone does not allow to discriminate $\mathcal{H}_{2}$ from
$\mathcal{H}_{3}$. The second hypothesis (\ie mutual modelling is primarily
supported by the capacity of the modellee to help the modellers to form a
correct image of herself), is questioned by study 3 where \C could be modelled
accurately by \A and not accurately by \B. The same hypothesis is however
supported by the results of study 2, and indirectly by study 4.

@@

In other words, these studies do not conclude that one hypothesis is right and the
others are wrong, and this is indeed the main contribution of this article. We
started from the idea that partner modelling is essentially an individual skill
and that we would therefore improve the quality of collaboration by providing 
awareness tools, as those tools would act as a kind of prosthesis for partner
modelling. The role and importance of this individual skill cannot be discarded:
everyone has experienced the pleasure of interacting with colleagues who answer precisely to the question
we asked them and even guess the reason why we asked this question. Conversely,
everyone also experienced the frustration of someone referring about a third
person by his name, say Mike Smith, while knowing perfectly that there is no
chance that we know this Mike Smith. In fact, the role of this individual skill is
not denied by our studies; the novelty of this paper is to show, thanks to
symmetry values, that individual skills only account for a part of the accuracy
in mutual modelling.

We acknowledge that the difference between these hypotheses is rather
theoretical since the  process of modelling one's partner $\mathcal{H}_{1}$ and
the process of helping one's partner to model oneself $\mathcal{H}_{2}$ are
mediated by verbal interactions in the team. One could hardly imagine someone
managing a very accurate modelling despite the very bad quality of interactions.
There is a bidirectional causal link between the accuracy of mutual modelling
and the quality of the interactions. This rather artificial distinction does
however participate to the more fundamental discussion of the role of individual
and social mechanisms in human cognition. In the field of social cognition, it
is a common place to state that ``the whole is greater than the sum of the
parts''.  This refers to the emergence of team properties than cannot be reduced
to the set of individual contributions. Our paper illustrates this emergence by
showing the symmetry of mutual modelling.

These conclusions must be presented with multiple disclaimers. First, they
heavily rely on correlations; hence we cannot identify causal links. Second, we
faced hard methodological issues. Providing learners with on-task questionnaires
introduces a bias: they will pay more attention to their partners in the
remaining time. Providing them with ``after-task'' questionnaires implies
mnemonic and rationalization biases. The nature of mutual modelling implies
methodological challenges that call for new measurement methods. We have
promising results for using eye tracking methods to address this challenge.
Third, our results emerge from a post-hoc reinterpretation of studies that
addressed different research questions (media richness, awareness tools,
scripts,...). This diversity makes our results difficult to integrate, partly
contradictory. Nonetheless, this diversity also incidentally provides some
generalizability: mutual modelling has been investigated in different contexts
(virtual space versus real space), with different groups sizes (pairs and
triads) and different tasks.

Finally, let us repeat that the expressions we used, such as \gmodel{A}{B}, do
not imply we have a mechanical view of modelling. It's just a language for
referring to different components of modelling each other in teams.  This simple
notation emerged from the need to compare five mutual modelling studies. It
could however pave the road for elaborating a systematic agenda for research on
mutual modelling.

>>>>

Collaborative learning has often been associated with the construction of  a
shared   understanding of the situation at hand. The psycholinguistics
mechanisms at work while establishing common grounds are the object of
scientific controversy. We postulate that collaborative tasks require some
level of mutual modelling, \ie that each partner needs some model of what
the other partners know/want/intend at a given time.  We use the term ``some
model'' to stress the fact that this model is not necessarily detailed or
complete, but that we acquire some representations of the persons we
interact with. The question we address is: Does the quality of the partner
model depend upon the modeler's ability to represent his or her partner?
Upon the modelee's ability to make his state clear to the modeler? Or rather upon
the quality of their interactions? We address this question by comparing
the respective accuracies of the models built by different team
members. We report on 5 experiments on collaborative problem solving
or collaborative learning that vary in terms of tasks (how important it is
to build an accurate model) and settings (how difficult it is to build an
accurate model). In 4 studies, the accuracy of the model that \A built about
\B was correlated with the accuracy of the model that \B built about \A,  which
seems to imply that the quality of interactions matters more than individual
abilities when building mutual models.  However, these findings do not
rule out the fact that individual abilities also participate in the quality of
modelling process.

@@

From its inception, Computer Supported Collaborative Learning (CSCL)
research has been following the suggestion by \citet{roschelle1995construction}
that collaborative learning has something to do with the process of constructing
and maintaining a \emph{shared understanding} of the task at hand. Building a
shared/mutual understanding refers to the upper class of collaborative learning
situations, those in which students should build upon each other's understanding
to refine their own understanding: what is expected to produce learning is not
the mere fact that two students build the same understanding but the cognitive
effort they have to engage to build this shared
understanding~\citep{schwartz1995emergence}. This effort can be observed by the
frequency of rich interactions, \ie interactions whose occurrence has been
related to learning: (self-) explanations in cognitive
science~\citep{chi1989self, webb1991task}, conflict resolution in
socio-cognitive theories~\citep{doise1975social} and mutual
regulation~\citep{blaye1995collaborative} in a Vygostkian perspective. The
construction of a shared understanding has been investigated for several decades
in psycholinguistics, under the  notion of
\emph{grounding}~\citep{clark1986referring}. However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned in the learning
sciences. Grounding mechanisms are appropriate to explain conversational
events, such as referential failures in short dialogue episodes, but they hardly
predict deeper phenomena such as \emph{conceptual change} (\ie the acquisition,
acceptation and integration of a new belief into one's mental model) over longer
sessions~\citep{dillenbourg2006sharing}. The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:

@@

In summary, the controversy around the cognitive depth of shared understanding
pertains to psycholinguistics, which investigates natural conversations. The
situation is different in collaborative learning; tasks are actually designed
for requiring a deeper negotiation of meaning.  The expertise of CSCL is to design
collaborative situations that create interdependence, avoid group think, and
allow students to detect any ``illusion of shared
understanding''~\citet{cherubini2005grounding}. Our question is hence not
anymore ``do peers build a shared
understanding ?'', but rather ``whenever peers have to build a shared understanding,
how do they achieve it ?''.  This question addresses the mechanisms of grounding, in
which the basic sequence is: make a proposition, detect misunderstanding and, if
any, repair them.  This paper focuses on the middle part, the detection of
misunderstandings.  Detecting a misunderstanding means that the emitter of a
message identifies a mismatch between his communicative intention and the way
his message is understood by his or her partner. Detecting one peer's
misunderstanding is investigated hereafter under the general umbrella of partner
modelling.

@@

    \item $Model(A,B, ``urgent")$: what does \A know about \B's understanding of \A's
        last utterance: does ``urgent'' mean now, soon or ``not too late''?

@@

This mutuality allows us to address a fundamental question: \emph{does the
quality of partner modelling depend upon the cognitive skills of each partner
(some people being better in perceiving other's states) or does it result from
the quality of interactions among them ?} Behind his question, the reader may
perceive the long lasting debate between tenets of, respectively, the individual
and social views of human cognition.  The simple hypothesis is that when
individuals are randomly paired for an experiment, there is no reason for which
their individual cognitive skills would correlate. Therefore, if it occurs that
the quality of the $Model(A,B)$ is correlated with $Model(B,A)$, one may infer
that this quality depends upon what \A and \B have built together while
interacting.

To answer this question, we went back to five previous studies that addressed
various other research questions, but in which the correlation between the
quality of these two models could be computed.  The studies we report do not hence
constitute a clean sequence of experiments to investigate mutual modelling but
the ad-hoc revisiting of previous experiments to explore a question that we had
then neglected. Some of these studies are about collaborative learning while
others are only collaborative problem solving, but the latter rely on rich tasks
that are similar to those we use in CSCL.

@@

Natural language becomes cumbersome when describing things such as ``the model
that \A builds about the model that \B builds about A''. Therefore, to define
hypotheses and report on experiments on mutual modelling, we use  the notation
\model{A}{B}{X} to denote ``\A knows that \B knows X''. This  notation is not
proposed as a formal theory of mutual modelling but as useful simplification for
communicating about mutual modelling. This notation does not mean \A has an
explicit, monolithic representation of \B: it must be understood as an
abstraction referring to complex socio-cognitive processes.  As explained in the
previous section, the model built by \A can be fragmented, multi-dimensional, etc. This notation
is neither presented as a computational model of mutual modelling, nor as some
universal formalism; its usefulness is internal to this paper.  Besides, we
refer to the \emph{degree of accuracy} of the model as \Model{A}{B}{X}. We
discuss in the next section the methodological difficulty in measuring this
accuracy.

@@

    \item Before any specific grounding action, there is generally a non-null
        probability that $X$ is mutually understood by \A and \B (\eg $X$ is
        part of \A's and \B's cultures, it is manifest to co-present subjects
        or simply there is not much space for misunderstanding or disagreement
        about $X$). We simply could not collaborate without a certain level of
        initial grounds. We denote the theoretical accuracy of \emph{initial
        grounds} with $\mathcal{M}^{\circ}_{t_0}(A,B,X)$.
        \vspace{1em}

@@

We report on five studies conducted  by different researchers in different contexts
between 2000 and 2015. They do not form a consistent research strategy but
the fact that some trends emerge despite their diversity constitutes the
richness of this line of work.

@@

How to interpret a correlation of 0.38 between \gModel{A}{B} and \gModel{B}{A}?
It is significant, which supports $\mathcal{H}_{2}$ and $\mathcal{H}_{3}$. It is
nonetheless far from 1, which implies we cannot discard the individual modelling
skills. We collect more evidence in the next studies.

@@

\paragraph{Rectangle question} We cannot address the relationship $\Delta_4$
between self and social accuracy here because we do not have an estimation of
self-accuracy: subjects describe their own emotions but we have no way to check if
they are correct. By measuring \gModel{A}{B} on  18 emotional labels,
we can however have a glimpse about $\Delta_5$: how \Model{A}{B}{X} varies
according to X.  Fig.~\ref{study3:deg_m_values} shows the range of modelling
errors: the difference between \gmodel{A}{B} and \gmodel{B}{B}, on a scale of 7,
is 0.3 in average for the emotion \emph{discasted}, and up to 1.9 for the
emotion \emph{calm}. This is probably specific to the variety of scales  ($SD=
0.5$ for \emph{discated} versus $SD=0.7$ for \emph{calm}). Our point is not to
interpret this too far, but to show that there are large variations even within one
area  (perceiving emotions). These variations still question the existence of
a general aptitude to model others ( $\mathcal{H}_{1}$).

@@

\paragraph{Grounding criterion} A linear regression revealed a positive relation
between \gModel{A}{B} and the learning gain of the pair $\{A, B\}$ (calculated
as the average of the individual learning gains): $\beta= 0.401, p < 0.001, r2adj.
= 0.15$, large effect. This relationship was not significant in the previous
experiment which was conducted with the same task (using the same task
should produce the same grounding criterion). This is probably due to the fact that
\gModel{A}{B} was influenced by the KAT.

@@

\paragraph{Symmetry question} We did not find an intra-pair correlation ($ICC =
0.05, ns$), which does not support $\mathcal{H}_{2}$ and $\mathcal{H}_{3}$. In
addition, the KAT supports the processing of modelling at the first level
($\mathcal{H}_{1}$) and not at the second level ($\mathcal{H}_{2}$). Hence the
fact that the KAT enhances  \gModel{A}{B}  brings additional supports to
$\mathcal{H}_{1}$. However, the analysis of verbal interactions finds elements
that actually supports the three hypotheses.

@@

We found a symmetry of the mutual models on 4 studies out of 5. In study 4, this
only applies to the control group (having read the same text before), which
corresponds to the situation of the 3 first studies. Evidencing this symmetry
constitutes \textit{per se} an interesting result as we are not aware of earlier
studies that have established this relationship.
Still, the symmetry alone does not allow to discriminate $\mathcal{H}_{2}$ from
$\mathcal{H}_{3}$. The second hypothesis (\ie mutual modelling is primarily
supported by the capacity of the modellee to help the modellers to form a
correct image of herself), is questioned by study 3 where \C could be modelled
accurately by \A and not accurately by \B. The same hypothesis is however
supported by the results of study 2, and indirectly by study 4.

@@

In other words, these studies do not conclude that one hypothesis is right and the
others are wrong, and this is indeed the main contribution of this article. We
started from the idea that partner modelling is essentially an individual skill
and that we would therefore improve the quality of collaboration by providing 
awareness tools, as those tools would act as a kind of prosthesis for partner
modelling. The role and importance of this individual skill cannot be discarded:
everyone has experienced the pleasure of interacting with colleagues who answer precisely to the question
we asked them and even guess the reason why we asked this question. Conversely,
everyone also experienced the frustration of someone referring to a third
person by his name, say Mike Smith, while knowing perfectly that there is no
chance that we know this Mike Smith. In fact, the role of this individual skill is
not denied by our studies; the novelty of this paper is to show, thanks to
symmetry values, that individual skills only account for a part of the accuracy
in mutual modelling.

We acknowledge that the difference between these hypotheses is rather
theoretical since the  process of modelling one's partner $\mathcal{H}_{1}$ and
the process of helping one's partner to model oneself $\mathcal{H}_{2}$ are
mediated by verbal interactions in the team. One could hardly imagine someone
managing a very accurate modelling despite the very bad quality of interactions.
There is a bidirectional causal link between the accuracy of mutual modelling
and the quality of the interactions. This rather artificial distinction does
however participate in the more fundamental discussion of the role of individual
and social mechanisms in human cognition. In the field of social cognition, it
is a commonplace to state that ``the whole is greater than the sum of the
parts''.  This refers to the emergence of team properties than cannot be reduced
to the set of individual contributions. Our paper illustrates this emergence by
showing the symmetry of mutual modelling.

These conclusions must be presented with multiple disclaimers. First, they
heavily rely on correlations; hence we cannot identify causal links. Second, we
faced hard methodological issues. Providing learners with on-task questionnaires
introduces a bias: they will pay more attention to their partners in the
remaining time. Providing them with ``after-task'' questionnaires implies
mnemonic and rationalization biases. The nature of mutual modelling implies
methodological challenges that call for new measurement methods. We have
promising results for using eye tracking methods to address this challenge.
Third, our results emerge from a post-hoc reinterpretation of studies that
addressed different research questions (media richness, awareness tools,
scripts,...). This diversity makes our results difficult to integrate, partly
and contradictory. Nonetheless, this diversity also incidentally provides some
generalizability: mutual modelling has been investigated in different contexts
(virtual space versus real space), with different groups sizes (pairs and
triads) and different tasks.

Finally, let us repeat that the expressions we used, such as \gmodel{A}{B}, do
not imply we have a mechanical view of modelling. The notation is just a language for
referring to different components of modelling each other in teams.  This simple
notation emerged from the need to compare five mutual modelling studies. It
could however pave the road for elaborating a systematic agenda for research on
mutual modelling.
